---
title: "fmi2: introduction"
author: "Joona Lehtomäki (revised by Pedro J. Aphalo)"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
vignette: >
  %\VignetteIndexEntry{fmi2 introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

`fmi2` is not yet in CRAN, so you'll need to install it directly from GitHub. 
While you're at it, make sure you also install all the packages below as we'll
be using them in this tutorial.

```{r installs, eval=FALSE}
install.packages(c("DT", "ggplot2", "leaflet", "remotes", "sf", "tidyverse"))

remotes::install_github("ropengov/fmi2")
remotes::install_github("ropensci/skimr")
```

```{r load-libraries, message=FALSE, warning=FALSE}
library(fmi2)
library(ggplot2)
library(leaflet)
library(sf)
library(skimr)
```

# Getting started

You can retrieve weather state observation data with various temporal 
resolutions using `fmi2`. First thing you need to know is of course which 
location exactly you want to get the data for.

## Finding the station of interest

The FMI API provides multiple ways of defining the spatial query area:

+ **Bounding box** given by coordindates and defining an area
+ **Place name** for which to provide data
+ **FMISID** numeric FMI observation station identifier
+ **GEOID** numeric geoid of the location
+ **WMO** code of the location

We'll start off by using the unique **FMISID** identifiers assigned to the [FMI
observation stations](https://en.ilmatieteenlaitos.fi/observation-stations). The
online table is also available in `fmi2` using the function `fmi_stations()`:

```{r show-stations}
station_data <- fmi2::fmi_stations() 

station_data
```

Some examples of how to find the station of interest. First by partial name match and partial match to available data Group. We for a station in Helsinki with precipitation data.

```{r find-by-partial-name-match}
station_data %>% 
  dplyr::filter(grepl("Helsinki", Name) & grepl("Precipitation", Groups))
```

Stations with WMO code assigned.

```{r find-has-WMO}
station_data %>% 
  dplyr::filter(!is.na(WMO))
```

Find longest series.

```{r find-longest}
station_data %>% 
  dplyr::arrange(Started) %>% 
  dplyr::slice(1)
```

Arrange stations from North to South.

```{r arrange-N2S}
station_data %>% 
  dplyr::arrange(-Lat)
```

The three stations at highest elevation.

```{r find-by-elevation}
station_data %>% 
  dplyr::arrange(-Elevation) %>% 
  dplyr::slice(1:3)
```

Select by range of latitude.

```{r find-by-lat-range}
station_data %>% 
  dplyr::filter(dplyr::between(Lat, 60, 60.1))
```

## Displaying the location on a map

We're going to pick "Hanko Tulliniemi" as an example here and use its FMISID
(100946) to retrieve the data. As you can see from the table above, it also 
provides the Lat Lon (geographical) coordinates for the observation station.
Before we get the actual data, let's visualize Hanko region.

```{r plot-hanko-map}
# Get data for Tulliniemi only
tulliniemi_station <- station_data %>% 
  dplyr::filter(FMISID == 100946)

# Plot on a map using leaflet
leaflet::leaflet(station_data) %>% 
  leaflet::setView(lng = tulliniemi_station$Lon, 
                   lat = tulliniemi_station$Lat, 
                   zoom = 11) %>% 
  leaflet::addTiles() %>%  
  leaflet::addMarkers(~Lon, ~Lat, popup = ~Name, label = ~as.character(FMISID))
```


# Getting daily weather observation data

Now that we know how the id for a specific observation station, we can proceed
to getting the actual data. `fmi2` provides several functions for retrieving 
data with different variables and temporal resolution. We'll start with 
`obs_weather_daily()` which returns daily average observation data from a given 
location. Let's get the daily weather observation data for the first 6 monhts
of 2019:

```{r getting-data}
# Use Hanko Tulliniemi weather station FMISID
tulliniemi_data <- obs_weather_daily(starttime = "2019-01-01",
                                     endtime = "2019-06-30",
                                     fmisid = 100946)
```

In total, the function returned `r nrow(tulliniemi_data)` observations. You can also
note the following:

```{r sf-class}
class(tulliniemi_data)
```

which means that the data returned by `obs_weather_daily()` is a spatial `sf`
object with the `geometry` column storing the geographical information of the
weather station. We'll come back to this later. Now we are interested in what 
*kind* of data did we actually get? Let's find out:

```{r}
unique(tulliniemi_data$variable)
```

So there are six variables with their corresponding values. `fmi2` provides a
helper function `describe_variables()` that can be useful in finding out more
about the variables:

```{r describe-variables}
fmi2::describe_variables(tulliniemi_data$variable)
```
`obs_weather_daily()` returns data  in so called  long (or melted) format 
meaning that all variable (i.e. parameter) names are  contained in column 
`variable` and corresponding values in `value` column. You can transform the 
data into a wide format using `tidyr`:

```{r spread-data}
wide_data <- tulliniemi_data %>% 
  tidyr::spread(variable, value) %>% 
  # Let's convert the sf object into a regular tibble
  sf::st_set_geometry(NULL) %>%
  tibble::as_tibble()

class(wide_data)
wide_data
```

Looks like there aren't many observations in the data for `rrday`, `snow` or `TG_PT12H_min`. 
Let's have a closer look at the data stored in `wide_data`:

```{r skim-data, results='asis'}
knitr::kable(skimr::skim(wide_data))
```

Seems like the above mentioned variables indeed don't have data between the
defined days. Let's get the same data from a couple of other observation 
stations around finland. Note that this time we're using place name instead of
a FMISID.

```{r multiple stations, warning=FALSE}
oulu_data <- obs_weather_daily(starttime = "2019-01-01",
                               endtime = "2019-06-30",
                               place = "Oulu")
nuorgam_data <- obs_weather_daily(starttime = "2019-01-01",
                                  endtime = "2019-06-30",
                                  place = "Nuorgam")
# Add location name to each data set and combine them
oulu_data$location <- "Oulu"
nuorgam_data$location <- "Nuorgam"
tulliniemi_data$location <- "Hanko"

all_data <- rbind(tulliniemi_data, oulu_data, nuorgam_data)
# Factorize location and make order explicit
all_data <- all_data %>% 
  dplyr::mutate(location = factor(location, 
                                  levels = c("Nuorgam", "Oulu", "Hanko"),
                                  ordered = TRUE))
```

Let's plot the daily temperature data for the different locations:

```{r plot-data}
all_data %>% 
  dplyr::filter(variable %in% c("tday", "tmax", "tmin")) %>% 
  ggplot(aes(x = time, y = value, color = variable)) + 
  geom_line() + 
  scale_color_manual(name = "",
                     values = c(tmax = "red", tday = "black", tmin = "blue"),
                     breaks = c("tmax", "tday", "tmin"),
                     labels = c(tmax = expression(t[max]), 
                                tday = expression(t[mean]), 
                                tmin = expression(t[min]))) +
  facet_wrap(~ location, ncol=1) + 
  ylab("Temperature, daily (C)\n") +
  xlab("\nDate") + theme_minimal()
```

# Getting hourly weather observation data

In addition to daily values, it is also possible to retrieve weather observation
data with finer temporal resolution, such as hourly data, using the function
`obs_weather_hourly()`. The data retrieved this has slightly different content
as compared to the daily data:

```{r hourly-data}
# Get the hourly observations for the first day of February 2019 in Hanko Tulliniemi
tulliniemi_data <- fmi2::obs_weather_hourly(starttime = "2019-02-01",
                                            endtime = "2019-02-02",
                                            fmisid = 100946)
```

Again, let's first have a look at what we actually got:

```{r descrive-variables-2}
fmi2::describe_variables(tulliniemi_data$variable)
```


# Getting minute sun radiation observation data

Now that we know how the id for a specific observation station, we can proceed
to getting the actual data. `fmi2` provides several functions for retrieving 
data with different variables and temporal resolution. We'll start with 
`obs_weather_daily()` which returns daily average observation data from a given 
location. Let's get the daily weather observation data for the first 6 monhts
of 2019:

```{r getting-sun-data}
# Use Kumpula station FMISID
kumpula_sun_data <- obs_sunrad_minute(starttime = "2019-01-01",
                                     endtime = "2019-01-02",
                                     fmisid = 101004)
```

In total, the function returned `r nrow(tulliniemi_data)` observations. You can also
note the following:

```{r sf-class-kumpula}
class(kumpula_sun_data)
```

which means that the data returned by `obs_weather_daily()` is a spatial `sf`
object with the `geometry` column storing the geographical information of the
weather station. We'll come back to this later. Now we are interested in what 
*kind* of data did we actually get? Let's find out:

```{r}
unique(kumpula_sun_data$variable)
```

```{r descrive-variables-sun}
fmi2::describe_variables(kumpula_sun_data$variable)
```
```{r}
wide_sun_data <- kumpula_sun_data %>% 
  tidyr::spread(variable, value) %>% 
  # Let's convert the sf object into a regular tibble
  sf::st_set_geometry(NULL) %>%
  tibble::as_tibble()

class(wide_sun_data)
wide_sun_data
```

```{r}
ggplot(wide_sun_data, aes(time, ifelse(DIFF_1MIN > 2, DIFF_1MIN / (DIR_1MIN + DIFF_1MIN), NA))) +
  geom_line()
```

```{r}
ggplot(wide_sun_data, aes(time, GLOB_1MIN)) +
  geom_line()
```

# Advanced

For finding stored queries available from the server we first need to dowsload 
the whole list of stored queries.

```{r}
api_obj <- fmi_api("DescribeStoredQueries")
nodes <- api_obj$content
unlist(purrr::map(nodes, xml2::xml_attrs)) -> stored.query.ids
```

If we are not sure what is the exact id, we can search partial matches. For
example `"radiation"` is used for sun radiation.

```{r}
stored.query.ids[grepl("radiation", stored.query.ids) & 
                   grepl("fmi", stored.query.ids)]
```
```{r}
list_parameters("fmi::observations::radiation::simple")
```

```{r}
list_parameters("fmi::observations::radiation::timevaluepair")
```

```{r}
stored.query.ids[grepl("stuk::", stored.query.ids)]
```
